---
title: "Nick’s Write-up"
author: "Nicholas Sclafani"
date: "5/12/2018"
output: pdf_document
latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

```{r}
load("TrainData.Rdata")
library(tidyverse)
```

## Unsupervised Learning



### PCA

While it is difficult to visualize the structure of 248-dimensional data, we can perform PCA to gather a better sense for the how our data looks along the most important PC directions.

```{r}
x <- data.matrix(trainX)
pc <- prcomp(x)
pc_df <- data.frame(pc$x[,1:10], y = as.factor(trainY))
ggplot(pc_df, aes(x = PC1, y = PC2, color = y)) + geom_point()
```

Looking at just the first two principle components we can already some structure in the data with Class 0 points for the most part clustered around the origin and Class 1 points for the most part are negative in PC 1 and around the origin in PC 2.

We can use the first 10 PC direction to explain __ % of the data.

### K-Means

```{r}
# PCA + K-Means
k2 <- kmeans(pc$x[,1:10], centers = 2, iter.max = 100, nstart = 10, 
                algorithm = "Lloyd")
x_df <- data.frame(pc$x[,1:10], cluster1 = as.factor((k2$cluster - 2)*-1),
                   cluster2 = as.factor(k2$cluster - 1))
x_df$y <- as.factor(trainY)

if (sum(x_df$cluster1 != trainY) < sum(x_df$cluster2 != trainY)) {
  x_df$cluster = x_df$cluster1
} else {
  x_df$cluster = x_df$cluster2
}

error_rate = sum(x_df$cluster != trainY) / 133

x_df <- x_df %>% mutate(correct = 
                          case_when(
                            (cluster !=  y) ~ "x",
                            TRUE ~ ""))
ggplot(x_df, aes(x = PC1, y = PC2)) + geom_point(aes(color = cluster)) + 
  geom_text(aes(label = correct))
```

We can see even with the just the first 2 PC scores that the mistakes were all made at points that should be in Class 1 but that lie near the origin where most of Class 0 is clustered. The error rate is:
```{r}
error_rate
```

We noticed that the error rate without dimension reduction is about the same, but with more severe dimension reduction, i.e. using only the first two principal components, there is a slight increase in error.


## Supervised Learning

### K-NN

For K-NN, our goal is to determine the number of neighbors, K, that best classifies our data. In order to do this, we ran 5-fold cross-validaiton. The smallest error was consistentily found with K = 1, K = 2, K = 3, and K = 4. While our first instict was to choose K = 3 to prevent noise, we found that K = 1 consistently had the smallest standard deviation in its error results, and this a 1 Nearest Neighbor model is our optimal choice for a K-NN model.
```{r}
set.seed(99)
library(class)
# without dimension reduction
folds <- sample(rep(1:5, 27), replace = FALSE)[1:133]
misclass <- matrix(rep(-1, 100), ncol = 5)
for (m in 1:20) {
  for (f in 1:5) {
    xtr <- trainX[which(folds != f),]
    xte <- trainX[which(folds == f),] 
    ytr <- trainY[which(folds != f)]
    yte <- trainY[which(folds == f)]
    
    yhat <- knn(xtr, xte, ytr, k = m)
    misclass[m, f] = sum(yhat != yte) / length(yhat)
  }
}
c <- apply(misclass, 1, mean)
sd <- apply(misclass, 1, sd)
knn_df <- data.frame(c, sd, k = 1:20)
ggplot(knn_df, aes(x = k, y = c)) + geom_point() + 
  geom_errorbar(aes(ymin=c-sd, ymax=c+sd)) + 
  labs(x = "K", y = "Misclassification Rate")
```
